---
title: "Team Passing Network Visualisation"
author: "Jocelyn Mara, University of Canberra"
date: "2022-04-01"
---

## Set up

```{r}
# Load the packages we will be using:
library(tidyverse)
library(ggsoccer)
```

```{r}
# Source the pitch dimensions from the `soccer-pitch-dimensions_statsbomb.R` code file
# assumes the file is in the 'code' directory
# which is in the current working directory
source("code/soccer-pitch-dimensions_statsbomb.R")
```

## Get StatsBomb data

```{r}
# Get the URL for the match
# store the url in an object
url <- "https://raw.githubusercontent.com/statsbomb/open-data/master/data/events/19748.json"
```

```{r}
# store path and file name for where to save the data
file_name <- "data/chelsea_liverpool_event-data.json"

# download the data if the file doesn't already exist
if (!file.exists(file_name)){
  download.file(url, destfile = file_name)
}
```

## Read data into R

```{r}
# Read the data into R using the `fromJSON` function from the **jsonlite** package:
library(jsonlite)
event_data <- fromJSON(txt = file_name)
```

## Finding substitute times

```{r}
# We want to create a network map for just starting players
# So we need to subset data from match times prior to the first substitute
event_data %>%
  filter(team$name == "Chelsea FCW",
         type$name == "Substitution") %>%
  select(period, timestamp, minute, second)
```

```{r}
# Find the time of the first substitution for Chelsea
sub_time <- event_data %>%
  filter(team$name == "Chelsea FCW",
         type$name == "Substitution") %>%
  summarise(sub_time = min(minute)) %>%
  unlist %>%
  unname

sub_time
```

## Taking a data subset

```{r}
# Take a subset of the data with the following conditions:
event_data_subset <- event_data %>%
        # keep just chelsea's data
  filter(team$name == "Chelsea FCW", 
         # keep data prior to the first sub
         minute < sub_time, 
         # keep just these event types
         type$name %in% c("Pass", "Ball Receipt*", "Ball Recovery", "Shot", 
                          "Dispossessed", "Interception", "Clearance", 
                          "Dribble", "Shot", "Goal Keeper", "Miscontrol", 
                          "Error"))
```

## Player involvements for nodes

- The size of the nodes is based on the number of involvements per player
- The location of the nodes is based on each player's average location

```{r}
nodes <- event_data_subset %>%
  # unnest the location variable
  unnest_wider(location) %>%
  # fix the new variable names  
  rename("location_x" = ...1,
         "location_y" = ...2) %>%
  # flip the origin to be bottom left corner
  mutate(location_y = abs(location_y - 80)) %>%
  # calculate average x,y location and n events for each player
  group_by("player" = player$name) %>% 
  summarise(x = mean(location_x, na.rm = TRUE), 
            y = mean(location_y, na.rm = TRUE), 
            n_events = n())
```

## Nodes

```{r}
ggplot(data = nodes) +
  geom_point(aes(x = x, y = y, size = n_events, colour = n_events))
```

## Edge

- To create the edges, we need to calculate how many times each player passed to every other player
- To start, we will take a subset of just the passing data

```{r}
# filter pass data
pass_data <- event_data_subset %>%
  filter(type$name == "Pass") # note the use of $
```

## Select variables

- Store just variables of interest in their own objects:

```{r}
passing_player <- pass_data$player$name
receiving_player <- pass_data$pass$recipient$name
type <- pass_data$pass$type$name
outcome <- pass_data$pass$outcome$name
```

- Combine variables into a single *tibble* (tidyverse dataframe)

```{r}
pass_data <- tibble(passing_player,
                    receiving_player,
                    type,
                    outcome)
```

## Fix NA values

- Passes that were "completed" are stored as `NA` in the `outcome` variable
- "Regular" passes are also stored as `NA` in the `type` variable

```{r}
pass_data <- pass_data %>%
  # replace na values in the type and outcome variables
  replace_na(replace = list(type = "Regular", outcome = "Complete")) %>%
  # just keep passes that were incomplete or complete
  filter(outcome %in% c("Incomplete", "Complete"))
```


## Edges

- We are only interested in "completed" passes so we will just count these data

```{r}
edges <- pass_data %>%
  # calculate number of times each player passed to every other player
  group_by(passing_player, receiving_player) %>%
  summarise(n = n()) %>%
  # keep just complete cases
  na.omit()
```



## Player to player interactions

- Our n values are based on a given passer-receiver combination
- We want to calculate total interactions though, regardless of who was passer or receiver

```{r}
edges2 <- edges %>%
  group_by(player1 = pmin(passing_player, receiving_player),
           player2 = pmax(passing_player, receiving_player)) %>%
  summarise(n = sum(n))
```


## Minimum number of passes

- We may want to remove observations where there were < 2 passes between two players

```{r}
# keep just rows where n > 1
edges2 <- edges2 %>%
  filter(n > 1)
```

## Edges and nodes  

- Join the node data with the edge data

```{r}
network <- left_join(edges2, nodes,
                   by = c("player1" = "player"))

network <- left_join(network, nodes, 
                     by = c("player2" = "player"))
```

- Check the result to understand what has happened here..

## Rename variables  

- Rename the variables for player1 (`.x`) and player2 (`.y`):

```{r}
network <- network %>%
  # rename has the format new = old
  rename(player1_x = x.x, player1_y = y.x, player1_n = n_events.x,
         player2_x = x.y, player2_y = y.y, player2_n = n_events.y)
```

## Network nodes  

```{r}
ggplot(data = network) + 
  # add the points for player1
  geom_point(aes(x = player1_x, y = player1_y, 
                 size = player1_n, colour = player1_n)) +
  # add the points for player 2
  geom_point(aes(x = player2_x, y = player2_y, 
                 size = player2_n, colour = player2_n)) +
  # rename the axis and legend titles
  labs(x = "", y = "", colour = "n", size = "n")
```


## Network edges  

- Add the edges to the plot using `geom_segment`

```{r}
ggplot(data = network) + 
  
  # add the edges based on start and end locations for player1 and player2
  geom_segment(aes(x = player1_x, y = player1_y, 
                   xend = player2_x, yend = player2_y,
                   # set size and transparency based on number of involvements
                   # between each player
                   size = n, alpha = n),
               colour = "darkgrey") +
  
  # add player 1 nodes
  geom_point(aes(x = player1_x, y = player1_y, 
                 size = player1_n, colour = player1_n)) +
  # add player 2 nodes
  geom_point(aes(x = player2_x, y = player2_y, 
                 size = player2_n, colour = player2_n)) +
  
  # add names of player1 as text
  geom_text(aes(label = player1, x = player1_x, y = player1_y)) +
  
  # add names of player2 as text
  geom_text(aes(label = player2, x = player2_x, y = player2_y)) +
  
  # remove the legend
  theme(legend.position = "none")
```


## Adding the pitch

```{r}
ggplot(data = network) + 
  
  # add the pitch as a backdrop to the network plot
  annotate_pitch(dimensions = pitch_statsbomb_custom) +
  
  geom_segment(aes(x = player1_x, y = player1_y, 
                   xend = player2_x, yend = player2_y,
                   size = n, alpha = n),
               colour = "darkgrey") +

  geom_point(aes(x = player1_x, y = player1_y, 
                 size = player1_n, colour = player1_n)) +
    
  geom_point(aes(x = player2_x, y = player2_y, 
                 size = player2_n, colour = player2_n)) +
  
  geom_text(aes(label = player1, x = player1_x, y = player1_y), size = 3) +
  
  geom_text(aes(label = player2, x = player2_x, y = player2_y), size = 3) +
  
  theme(legend.position = "none") +
  
  theme_pitch()
```
